Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-05-23 paper/main.tex
work on section 3

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra and quantify uncertainty for a particular cosmology.
Let input $X$ represent the $n$-vector of $\log_{10}(k)$ values.
In the previous section, we used lowercase $\bar{y}$ to denote 
the observed weighted average; now, we use $\bar{Y}$ to indicate the corresponding 
random variable within the statistical model.  $\bar{Y}$ is typically pre-processed
in a domain-specific manner, such as that defined in Section \ref{sec:data}.
Together, $X$ and $\bar{Y}$ comprise the training data for a particular cosmology.
We will reserve specific integration 
of expert knowledge for the CAMB and Mira-Titan suites, such as this pre-processing,
for Sections \ref{subsec:camb} and \ref{subsec:mira_fit}, respectively.

\subsection{Deep Gaussian Process Hierarchical Model}

Here, we introduce the Bayesian Gaussian Process (GP) model. We assume 
the matter power spectrum $S$ for a given cosmology follows a GP with mean $\mu$ 
and covariance matrix $\Sigma_S$. From the data, we obtain $\bar Y$, which is a 
noisy version of $S$.  Let $S$ represent the true (but unknown) underlying
power matter spectra from which $D$ was generated.

Therefore, we model $\bar Y|S$ as a GP with mean $S$ and block-diagonal covariance 
matrix $\Sigma_\varepsilon$:

\begin{align}
\bar Y|S &\sim GP\left(S,\Sigma_\varepsilon\right) \\
S &\sim GP\left(\mu, \Sigma_S(x,x')\right)
\end{align}

@@

\begin{align}
\Sigma_\ell^{i,j} = \tau_\ell^2  \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_\ell}} + 
  \frac{5d^2}{3\theta_\ell}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_\ell}}\right),\\
\Sigma_S^{i,j} = \tau_S^2  \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_S}} + 
  \frac{5d^2}{3\theta_S}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_S}}\right),
\end{align}

where $d=||x_i-x_j||^2$. From here, we estimate $\tau_\ell, \theta_\ell$ with maximum 
likelihood, and conditional on $\Sigma_\varepsilon$, we specify priors for $\tau_S, \theta_S$ 
following \cite{sauer2023active}: \textcolor{red}{Annie, please ensure the most up-to-date 
prior specs are shown above and here. Here is what I found from fit\$settings when using 
deepgp v1.1.3: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}

@@

@@

Considering the Mira-Titan data, some nonstationarity is expected (e.g., due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$). To account 
for this, we incorporate a latent layer $W$ within the hierarchical model to warp 
locations of $X$ and incorporate nonstationary dynamics of the matter power spectra. 
When $W$ is a GP, this creates a deep Gaussian process (DGP) model \citep{damianou2013deep}. 
Additional latent layers could be considered, although one is sufficient for this 
dataset \citep{dunlop2018deep}.

\begin{align}
\bar Y|S,W &\sim \mathcal{N}_n(S,\Sigma_\varepsilon) \\
S|W &\sim GP\left(\mu, \Sigma_S(w,w')\right) \\
W &\sim GP\left(0, \Sigma_W(x,x')\right)
\end{align}

>>>>

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra and quantify uncertainty for a particular cosmology.
Let input $X$ represent the vector with elements $x_i = \log_{10}(k_i)$ for $i=1,\dots,n$.
In the previous section, we used lowercase $\bar{y}$ to denote 
the observed weighted average; now, we use $\bar{Y}$ to indicate the corresponding 
random variable within the statistical model.  $\bar{Y}$ is typically pre-processed
in a domain-specific manner, such as that defined in Section \ref{sec:data}.
Together, $X$ and $\bar{Y}$ comprise the training data for a particular cosmology.
We will reserve specific integration 
of expert knowledge for the CAMB and Mira-Titan suites, such as this pre-processing,
for Sections \ref{subsec:camb} and \ref{subsec:mira_fit}, respectively.

\subsection{Model Specification}

Let $S$ represent the true (but unknown) underlying power matter spectra for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e. 
$\bar{Y}|S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2024dpc}, we find it
essential to account for the smoothly varying spatial dependence of the
low resolution runs (as visualized in Figure \ref{fig:plot_data}).  For
this, we embrace variations of the Mat\'ern kernel \citep{stein1999interpolation},
tailored to each scenario (more on this in later subsections).

We next place a zero-mean Gaussian process prior on $S$, i.e.,
$S\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(\cdot)\right)$, again with a 
spatial Mat\`ern kernel.  
\textcolor{red}{(@Steve - is a zero mean the right thing?  If not, what is the mean?)}
Standard practice would be to use locations $X$ as the inputs to kernel $\Sigma_S(\cdot)$.
Yet we have found that prior to lack sufficient flexibility for our motivating
application.  In the Mira-Titan setting, some nonstationarity is expected due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$. To account 
for this, we incorporate an additional latent Gaussian process, $W$, which accommodates
nonstationarity by warping $X$.  Ultimately, this yields the following hierarchical model:
\begin{equation}\label{eq:dgphm}
\begin{aligned}
\bar{Y}|S,W &\sim \mathrm{GP}(S,\Sigma_\varepsilon) \\
S|W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(W)\right) \\
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{aligned}
\end{equation}
The functional composition of GPs in the latter two lines forms a ``deep Gaussian process'' 
\citep{damianou2013deep,dunlop2018deep}. Additional latent layers could be considered, 
although we find one is sufficient.

\[
\Sigma_S(X)^{i,j} = \tau^2  \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta}} + 
  \frac{5d^2}{3\theta}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta}}\right)
\quad\textrm{for}\quad 
d=||x_i-x_j||^2
\]
Hyperparameters $\tau^2$ and $\theta$ govern the correlation scale and decay respectively.

@@

where. From here, we estimate $\tau_\ell, \theta_\ell$ with maximum 
likelihood, and conditional on $\Sigma_\varepsilon$, we specify priors for $\tau_S, \theta_S$ 
following \cite{sauer2023active}: \textcolor{red}{Annie, please ensure the most up-to-date 
prior specs are shown above and here. Here is what I found from fit\$settings when using 
deepgp v1.1.3: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}

@@

\subsection{Model Inference}

@@
